---
title: "README"
output: html_document
date: "2025-06-12"
---


# Purpose

The purpose of this project is to evaluate the interest rate dynamics in South Africa since 2000. This is for 2025 Data Science for Economics and Finance 871 project. 


# Data

```{r}

library(huxtable)
library(dplyr)
library(tidyr)
library(xtable)
library(here)
library(forecast)
library(vars)   
library(tseries)

# Importing Global Macro Dataset 
Macro  <- read_csv(here("data", "GMD.csv"))

# Filter the data to South Africa
MzansiMAcro <- Macro %>% filter(countryname == "South Africa")

```


# Visualization Section 


```{r}

# Plotting a graph that showed the years where Inflation has been outside the SARB target range since 2000 
MzansiMAcro %>% filter(year >=2000 & year<=2024) %>% mutate(outside_target = infl < 3 | infl > 6) %>% ggplot(aes(x = year, y = infl)) +
  geom_line(color = "grey60") +
  geom_point(aes(color = outside_target), size = 3) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 6, linetype = "dashed", color = "blue") +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red"), guide = "none") +
  labs(
    title = "Years with Inflation Outside SARB Target Range (3–6%)",
    x = "Year",
    y = "Inflation Rate (%)"
  ) +
  theme_minimal()

```

```{r}

 #Visualizing the inflation and interest rate trends since 2000

sa_rates_long5 <- MzansiMAcro %>% filter(year >= 2000 & year <= 2024)%>% pivot_longer(cols = c(strate, ltrate, infl),names_to = "rate_type", values_to = "rate")

 ggplot(sa_rates_long5, aes(x = year, y = rate, color = rate_type)) +
  geom_line(linewidth = 1.2) + scale_y_continuous(limits = c(-5, 25)) + scale_x_continuous(breaks = seq(2000, 2025, by = 5)) +
  labs(title = "Inflation and Interest Rate Dynamics in South Africa ",
       x = "Year", y = "Rate (%)",
       color = "Rate Type") +
  theme_minimal() 


```


## Statistical Tests

### Inflation


```{r Figure10, echo=FALSE, fig.align='center', fig.cap="Autocorrelation Function of Inflation \\label{Figure10}", fig.height=3, fig.width=6, warning=FALSE, dev='png'}

# Use clean_df data for the statistical modelling section 

# Note the 2024 data is dropped here because there is a NA in the long-term interest rate 
clean_df <- MzansiMAcro %>%
  filter(year >= 2000 & year <= 2024) %>%
  dplyr ::select(year , infl, short_rate = strate, long_rate = ltrate) %>%
  na.omit()


# For the inflation

# converting the datq to a time series object 
infl_ts <- ts(clean_df$infl, start = 2000, frequency = 1)
short_ts <- ts(clean_df$short_rate, start = 2000, frequency = 1)
long_ts <- ts(clean_df$long_rate, start = 2000, frequency = 1)

 # A autocorrelation test 
# ACF shows you how a lag period is directly or indirectly related to our current value
acf(infl_ts, main = "ACF of Inflation Rate")
# The ACF plot indicates that inflation shows significant autocorrelation for up to two lags, while the third and subsequent lags are statistically insignificant. This suggests that inflation possesses short-term memory, with its effects fading relatively quickly over time.





```


```{r Figure11, echo=FALSE, fig.align='center', fig.cap="Partial Autocorrelation Function of Inflation\\label{Figure11}", fig.height=3, fig.width=6, warning=FALSE, dev='png'}


# A partial autocorrelation test 
# PACF shows you how a lag period is directly related to our current value , this is done by controlling for the immediate lags 

pacf(infl_ts, main = "PACF of Inflation Rate")




```

```{r include=FALSE}

# ACF and PACF tests show some non-stationarity but ADF test shows stationarity 

adf.test(infl_ts)

```

```{r include=FALSE}
## Comparing ARIMA models to select the most appropriate for modelling inflation , given that ADF indicates no stationarity , and no clear cutoff after lag 1 that would indicate a Moving Average Structure 

library(forecast)
# Fit AR(1)
model_ar1 <- Arima(infl_ts, order = c(1,0,0))  
model_ar1
# Fit AR(2)
model_ar2 <- Arima(infl_ts, order = c(2,0,0))
model_ar2




```


### Short-Term Interest 

```{r Figure12, echo=FALSE, fig.align='center', fig.cap="Autocorrelation Function of Short-Term Interest Rate\\label{Figure12}", fig.height=3, fig.width=6, warning=FALSE, dev='png'}

# ACF For the short_rate 

acf(short_ts, main = "ACF of Short-Term Interest Rate")

```

```{r Figure13, echo=FALSE, fig.align='center', fig.cap=" Partial Autocorrelation Function of Short-Term Interest Rate\\label{Figure13}", fig.height=3, fig.width=6, warning=FALSE, dev='png'}

#PACF for short_rate 
pacf(short_ts, main = "PACF of Short-Term Interest Rate")
```



```{r include=FALSE}
#Augmented Dickey Fuller test , this is the formal test for stationarity 
adf.test(short_ts)
```



```{r include=FALSE}
# Taking the first difference, then check if the first difference is stationary 
diff_short <- diff(short_ts)

adf.test(diff_short)

```




### Long-Term Interest 

```{r Figure14, echo=FALSE, fig.align='center', fig.cap="Autocorrelation Function of Long-Term Interest Rate\\label{Figure14}", fig.height=3, fig.width=6, warning=FALSE, dev='png'}

# ACF For the long_rate 
acf(long_ts, main = "ACF of Long-Term Interest Rate")



```


```{r Figure15, echo=FALSE, fig.align='center', fig.cap="Partial Autocorrelation Function of Long-Term Interest Rate\\label{Figure15}", fig.height=3, fig.width=6, warning=FALSE, dev='png'}

# PACF for long_rate

pacf(long_ts, main = "PACF of Long-Term Interest Rate")

```


```{r include=FALSE}
# ADF test for long
adf.test(long_ts)

```

```{r include=FALSE}

# Take the first difference
diff_long <- diff(long_ts)
# Test for stationarity 
adf.test(diff_long)

# Take the second difference , first difference is non-stationary 
diff2_long <- diff(diff_long)
# Test for stationarity 
adf.test(diff2_long)


# Take the third difference , second difference is non-stationary 
diff3_long <- diff(diff2_long)
# Test for stationarity 
adf.test(diff3_long)


```



## Modelling 

The Autoregressive Model of order 2 for inflation is : 
\begin{equation}
\pi_t = c + \phi_1 \pi_{t-1} + \phi_2 \pi_{t-2} + \varepsilon_t
\end{equation}


```{r include=FALSE}

# Manually constructing the output based on the results of the ARIMA 
coefs_df <- data.frame(
  Term = c("ar1", "ar2", "intercept"),
  Estimate = c(0.5240, -0.4711, 5.2471),
  Std_Error = c(0.1744, 0.1694, 0.3815)
)

# Create the xtable
model_table <- xtable(coefs_df, 
                      caption = "AR(2) Model Coefficients for Inflation \\label{tab66}")




```


```{r Liu, results = 'asis'}

# Print AR(2) output
print(model_table,
      floating = TRUE,
      include.rownames = FALSE,
      table.placement = "H",
      caption.placement = "bottom",
      comment = FALSE)
```



\bigskip 

The following Vector Autoregressive Model (VAR(6)) was used to examine the dynamic relationship between inflation and the short-term interest rate:


\begin{equation}
\begin{bmatrix}
\pi_t \\
r_t
\end{bmatrix}
=
\mathbf{c}
+
\sum_{i=1}^{6}
\Phi_i
\begin{bmatrix}
\pi_{t-i} \\
r_{t-i}
\end{bmatrix}
+
\begin{bmatrix}
\varepsilon_{\pi,t} \\
\varepsilon_{r,t}
\end{bmatrix}
\end{equation}



```{r include=FALSE}

library(vars)        


# remove the first inflation observation so that the VAR model has the same number of rows 
# combine the inflation and short-term interest into one matrix
var_data <- cbind(inflation = clean_df$infl[-1], short_rate = diff(clean_df$short_rate))

# Check lag length
VARselect(var_data, lag.max = 10, type = "const")

# Estimate VAR model with chosen lag , VARselect informs that the optimal is 6 lags 
var_model <- VAR(var_data, p = 6, type = "const")
summary(var_model)


# Manually construct the coefficient table 
var_coef_table <- data.frame(
  Term = c("inflation.l1", "short_rate.l1", "inflation.l2", "short_rate.l2",
           "inflation.l3", "short_rate.l3", "inflation.l4", "short_rate.l4",
           "inflation.l5", "short_rate.l5", "inflation.l6", "short_rate.l6", "const"),
  Estimate = c(-0.02951, 0.47257, 0.03001, -0.45616, 
               0.24376, -0.67412, -0.72850, 0.42824,
               0.70444, -0.32386, -0.22653, -0.34392, -0.31766),
  Std_Error = c(0.51052, 0.47297, 0.50395, 0.42000,
                0.32861, 0.45561, 0.33486, 0.62009,
                0.42630, 0.62727, 0.42468, 0.39421, 5.02632),
  p_value = c(0.9548, 0.3211, 0.9538, 0.2895,
              0.4602, 0.1725, 0.0952, 0.5030,
              0.1377, 0.6107, 0.6022, 0.3893, 0.9513)
)

# Create and print the LaTeX table
VAR_out <- xtable(var_coef_table, 
                     caption = "VAR(6) Model Coefficient for the Short-Term Interest Rate \\label{tab67}")



```

```{r Mee , results = 'asis'}


# Print VAR xtable 
print(VAR_out, 
      floating = TRUE,
      include.rownames = FALSE, 
      table.placement = "H", 
       caption.placement = "bottom",
     comment = FALSE)

```



# Appendix 

```{r}


library(tidyverse)
library(lubridate)
library(huxtable)
library(dplyr)
library(tidyr)
library(xtable)
library(vars)

# Generating a table that describes the variables used in the paper 
var_table <- data.frame(
  Variable = c("infl", "strate", "ltrate", "cbrate"),
  Description = c("Inflation rate (%)",
                  "Short-term interest rate (%)",
                  "Long-term interest rate (%)",
                  "Central bank rate (repo rate, %)")
)

vva <- xtable(var_table, caption = "Variable Descriptions\\label{tab21}")


  print.xtable(vva,
             # tabular.environment = "longtable",
             floating = TRUE,
             table.placement = 'H',
             # scalebox = 0.3,
             comment = FALSE,
             caption.placement = 'bottom',
             include.rownames = FALSE
             )



```

```{r}

 

library(tidyverse)
library(lubridate)
library(huxtable)
library(dplyr)
library(tidyr)
library(xtable)
library(vars)
library(here)

# Generating a table showing inflation since 2000 

 infa <- MzansiMAcro  %>% mutate(year = as.integer(year)) %>%  filter(year >= 2000 & year <= 2024) %>% dplyr::select(Year = year, `Inflation Rate (%)` = infl) 


  thea<-xtable(infa , caption = "Annual Inflation Rate in South Africa (2000–2024) \\label{tab8}" )
  
  print.xtable(thea,
             # tabular.environment = "longtable",
             floating = TRUE,
             table.placement = 'H',
             # scalebox = 0.3,
             comment = FALSE,
             caption.placement = 'bottom',
             include.rownames = FALSE
             )
  
```

```{r}


library(tidyverse)
library(lubridate)
library(huxtable)
library(dplyr)
library(tidyr)
library(xtable)
library(vars)
library(here)

# Generating a table showing the inflation and repo rate post the 2008 peak of inflation

bahn_table <- MzansiMAcro %>%mutate(year = as.integer(year)) %>% 
  filter(year >= 2010 & year <= 2024) %>%
  dplyr::select(Year = year, `Inflation Rate (%)` = infl , `Repo Rate (%)` = cbrate)


haibo<-xtable(bahn_table , caption = "Annual Inflation and Repo Rate in South Africa (2000–2024) \\label{tab9}" )


print.xtable(haibo, 
             floating = TRUE,
             table.placement = 'H',
             comment = FALSE,
             caption.placement = 'bottom',
             include.rownames = FALSE)
```


# Methodology \label{Meth}

## Subsection

Ideally do not overuse subsections. It equates to bad writing.[^1]

[^1]: This is an example of a footnote by the way. Something that should also not be overused.

## Math section

Equations should be written as such:

\begin{align}
\beta = \sum_{i = 1}^{\infty}\frac{\alpha^2}{\sigma_{t-1}^2} \label{eq1} \\
\int_{x = 1}^{\infty}x_{i} = 1 \notag
\end{align}

If you would like to see the equations as you type in Rmarkdown, use \$ symbols instead (see this for yourself by adjusted the equation):

$$
\beta = \sum_{i = 1}^{\infty}\frac{\alpha^2}{\sigma_{t-1}^2} \\
\int_{x = 1}^{\infty}x_{i} = 1
$$

Note the equation above - this should appear in math in Rstudio. Writing nice math requires practice. Note I used a forward slashes to make a space in the equations. I can also align equations using **&**, and set to numbering only the first line. Now I will have to type \`\`begin equation'' which is a native \LaTeX command. Here follows a more complicated equation that you want aligned (using & sign) and numbered:

\begin{align}
    y_t &= c + B(L) y_{t-1} + e_t   \label{eq2}    \\ \notag
    e_t &= H_t^{1/2}  z_t ; \quad z_t \sim  N(0,I_N) \quad \& \quad H_t = D_tR_tD_t \\ \notag
        D_t^2 &= {\sigma_{1,t}, \dots, \sigma_{N,t}}   \\ \notag
        \sigma_{i,t}^2 &= \gamma_i+\kappa_{i,t}  v_{i, t-1}^2 +\eta_i  \sigma_{i, t-1}^2, \quad \forall i \\ \notag
        R_{t, i, j} &= {diag(Q_{t, i, j}}^{-1}) . Q_{t, i, j} . diag(Q_{t, i, j}^{-1})  \\ \notag
        Q_{t, i, j} &= (1-\alpha-\beta)  \bar{Q} + \alpha  z_t  z_t'  + \beta  Q_{t, i, j} \notag
\end{align}

Note that in \ref{eq2} I have aligned the equations by the equal signs. I also want only one tag, and I create spaces using \`\`quads''.

See if you can figure out how to do complex math using the two examples provided in \ref{eq1} and \ref{eq2}.